Initial spatial conditions in simulation models: the missing leg of sensitivity analyses?

### 2017-08-03 paper.tex
Update on Overleaf.

\corrauth{Corresponding author
address.}

\email{c.cottineau@ucl.ac.uk}

@@

\begin{abstract}
%\noindent
%\setlength{\parindent}{0pt}
Although simulation models of geographical systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain until their results are proven robust. Sensitivity analysis usually include the analysis of the effect of stochasticity on the variability of results, as well as the effects of small parameter changes. However, initial spatial conditions are usually taken for granted in geographical models, thus leaving completely unexplored the effect of spatial arrangements on the interaction of agents and of their interactions with the environment. In this contribution, we present a method to assess the effect of initial spatial conditions on simulation models, using a systematic generator controlled by meta-parameter to create density grids used in spatial simulation models. We show, with the example of two very classical agent-based models (Schelling's models of segregation and Sugarscape) that the effect of space in simulation is significant, and sometimes even larger than parameters themselves. We do so using high performance computing in a very simple and straightforward open-source workflow.
\end{abstract}

@@

% **************  MAIN BODY OF THE PAPER **************

@@

Simulation has been recognised and increasingly used by geographers to explore various geographical processes and problems within virtual laboratories \citep{Quesneletal2009} \comment{Batty 1978? James Doran 1970's : lire thèse de Seb ;) }. It appears as a very fruitful way to overcome the difficulty of analytic resolution of many spatial models developed in the past, and to explore the possible trajectories of social and ecological systems in space and time. The specificity of geographical models compared with models from other social sciences is usually regarded as the way geographers consider space and spatial interactions, driven by an explicit interest in the way space influences the outcomes of the model. Geographers are indeed concerned about understanding and modelling how space plays a role in social interactions and environmental processes, and whether its action is placed-based or place-neutral. We think simulation can become a very good tool to achieve this, provided that models include relevant spatial description and modelling, behavioural rules that take space into account, and provided model evaluation stresses the sensitivity of output variations to the way space is modelled. This paper aims to fill a methodological and conceptual gap, which is a systematic testing of the sensitivity of model's outcomes to initial spatial conditions. To demonstrate the genericity of our approach, we develop two applications from simulation models which are commonly used as case studies for comparing and aligning simulation models \citep{Axtelletal1996}: Schelling's model of segregation and Epstein and Axtell's Sugarscape model. Both models are defined at the intraurban level.

An urban system consists of a large number of social agents that interact with each other at a microscopic level in a system-time that evolves irreversibly creating temporal and cumulative effects. Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,Portugali2000}, and emergent properties at qualitatively differentiated scales~\citep{AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002,Batty2007}. In that frame, complexity is partially due to bifurcations \citep{Wilson1981} which are determinant in spatial systems~\citep{Wilson2002}.

In spatially explicit simulation models, the non-linearity of local interactions is therefore very likely to sublimate small perturbations in the initial spatial setting, making it difficult to interprete the resulting global structures. \comment[JR]{je suis pour garder le passage sur la non-ergodicité (commenté ci-après) ; ça vous parait trop abstrait ?}\comment[MLT]{Non, mais je pense que la seconde phrase pourrait être reformulée/développée pour mieux raccrocher avec le reste, qu'en penses-tu?} A crucial aspect of most spatio-temporal complex systems is their non-ergodicity~(\cite{pumain2012urban}) (the property that cross-sectional samples in space are not equivalent to samples in time to compute statistics such as averages), what witnesses generally strong spatio-temporal path-dependencies in their trajectories. Similar to what Gell-Mann calls \emph{frozen accidents} in any complex system~\cite{gell1995quark}, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system.

At initialisation, we then expect the spatial distribution of agents to influence results in the long run \cite{Castellanoetal2009}, as agents' rule of action itself might depend on the spatial structure of the environment: households can have different preferences with respect to the built-environment (\cite{SpielmanHarrison2014}), mechanisms of surrounding sensing are impacted by different distributions of density (\cite{Banos2012}) or the scale of modelled environments (\cite{LauriJaggi2003, FossettDietrich2009}), etc. The way modellers abstract space is therefore a central element of any spatially explicite simulation model. Although this may seem obvious, cities are not regular grids of isotropic densities. However, the uniform grid which represents space in most simulation models is potentially not enough to represent urban processes because density and accessibility have environmental, economic and social consequences. An intermediate and more meaningful way of abstracting space might thus be to consider, not the peculiarities of every city, but their broad density structures. In Europe for example, one can find broad types of density distributions (\cite{LeNechet2015}).

@@

In this paper, we suggest tackling sensitivity to spatial conditions by generating a variety of density grids to feed into simulation models at initialisation, and exploring the sensitivity of the model outcomes to the variation of meta-parameters (i.e. parameters used to generate initial spatial conditions). The purpose is two-fold: 1/ to test the robustness of simulation results to small variations of meta-parameters within a typical category of space (a monocentric case for example\comment[FL]{je ne trouve pas cela facile à comprendre car on n'a pas encore parlé des différentes formes de villes}) and 2/ to study the non-trivial effects of typical categories of spatial distribution (monocentric \textit{vs.} polycentric for example) on the results of a given model.

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Previous considerations of urban form in simulation models}

\begin{itemize}
\item spatial context
\begin{itemize}
\item extent
\item objects
\item shape
\end{itemize}

\item spatial "encoding"
\begin{itemize}
\item heterogeneity
\item algorithm of disaggregation
\item uncertainty
\end{itemize}

\item spatial effects
\begin{itemize}
\item topology
\item interaction (??)
\end{itemize}

\item spatial outputs
\begin{itemize}
\item aggregation
\item use of models
\end{itemize}
\end{itemize}
%\comment[FL]{Je pense que cette partie est vraiment utile, et devrait être étoffée, en procédant peut être à une typologie plus systématique des "initial condition" qui nous intéressent. Il y a déjà une différence entre l'hétérogénéité de l'espace (densités réparties mono ou poly, etc) et les différentes façons de représenter l'espace (grilles carrées, hexagones). Par ailleurs aussi des discussions sur l'incertitude de l'information (est ce que par exemple les emplois downscalés au carreau INSEE par Pivano 2016 c'est vraiment fiable ou très ?) et la précision de la grille}

The effect of the spatial configuration on area-based attributes of human behaviours has been largely discussed in geostatistics, meanly since the exposure of the Modifiable Areal Unit Problem (MAUP) \citep{Openshaw1984,FotheringhamWong1991}. Recently, \citet{Kwan2012} claims for a careful examination of what she coins the uncertain geographic context problem (UGCoP), that is of the spatial configuration of geographical units even if the size and delineation of the area are the same. Considerations of such issues in the geographic simulation model literature are rather scarce, but they have been some noticeable examples of papers attempting to evaluate the effects of the same set of mechanisms runned from different initial spatial conditions. Mainly, these approaches have been focusing on the impact of geo-localised input data accuracy, of the spatial system shape, precision, and boundaries, as well as of the impact of spatial heterogeneity.

\textbf{Geo-localised input data accuracy:} \citep{Thomasetal2017} show that data selection in LUTI model is interelated to the delineation of the spatial sytem boundaries and scale of analysis. They provide a few examples of how the use of Exploratory Spatial Data Analysis (ESDA) prior to simulation runs help avoiding measurement errors of model behaviour and outcomes.

\textbf{Spatial system shape, precision, and boundaries:} \citet{FlacheHegselmann2001} show that chances for random emergence of a stable cluster of similar agents in a Schelling-like model are highest in a rectangular grid and lowest in a hexagonal grid and that an irregular (Voronoi-diagram) city lattice structure favours migration stabilisation around decentralised clusters of similar agents. \citet{Banos2012} compares the behaviour of Schelling segregation model on city lattices formalized as either grid, random, scale-free and Sierpinski networks and concludes that the presence of cliques in graph-based urban structures favor segregationist behaviours. \citet{LeTexierCaruso2017} simulate an individual-based dynamic model where agents move and exchange coins across regions. Using a set of different theoretical spatial systems, they demonstrate the impact of the regularity and aggregation levels, or centrality/periphery effects, on spatial diffusion dynamics.

\textbf{Spatial heterogeneity:} \citet{StaufferSolomon2007} introduce asymetric interactions and empty residences in Schelling's model run on a large and regular lattice. They reveal conjoint and non-linear effects on the vacancy rates and tolerance levels on segregation patterns. \citet{Gauvinetal2010} run Schelling's segregation process in an open city-lattice to study how the variations in tolerance levels, vacancy rates and city attractiveness may create lines of vacancy lots between clusters of agents. They conclude on the functional role of vacancies, which allow weakly tolerant agents to live and be satisfied in a city environment they nevertheless perceive as hostile. \citet{HatnaBenenson2012} show that their model replications run on a 50x50 torus with 2\% of empty cells were not sensitive to the initial patterns (random and fully segregated distribution of agents).

\textbf{Combination:} \citet{Singhetal2009} show that the segregation patterns obtained by Schelling for certain tolerance values are strictly a small city phenomenon (8x8 city-lattice) and do not work for a larger spatial lattice (100x100), where segregation appears only for certain combinaisons of tolerance threshold and vacancy density values.

In this paper, we build on the contributions of these previous studies and go further by systematically measuring the impact of the city lattice on model behaviour so far only assessed through a small set of illustrative spatial structures. We illustrate the potential generalization of our approach by running two distinctive agent-based models: Schelling's model of residential segregation and Sugarscape.

@@

In this section, we detail the method developed to analyse the sensitivity of simulation models to initial spatial conditions. The general method workflow is illustrated in Figure \ref{fig:method}. In addition to the usual protocol (upper branch in figure \ref{fig:method}, which consists of running a model $\mu$ with various values of its parameters and relating these variations of values to the variations in the simulation results, we here introduce a spatial generator (lower branch in figure \ref{fig:method}), which itself is determined by input parameters and produces sets of spatial initial conditions. Initial spatial conditions are clustered to represent types of spaces ex-ante (for example: monocentric or polycentric density grids), and the sensitivity analysis of the model is now run against $\mu$ parameters as well as spatial parameters or spatial types. It allows the sensitivity analysis to produce qualitative conclusions regarding the influence of spatial distribution on the outputs of simulation models, alongside the classic variation of parameter values.

%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[htbp] \begin{center} 
\resizebox{0.9\textwidth}{!}{ 
	\includegraphics{figures/SchemaMeta_1.png}
} \caption{General workflow of our method} \label{fig:method} \end{center} \end{figure*} %
%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Constructing a spatial generator}

Our spatial generator applies an urban morphogenesis model (\cite{Batty2007}), which has been generalised, explored and calibrated (\cite{Raimbault2014}). In a nutshell, grids are generated through an iterative process which, starting from a void grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment \comment[FL]{à quoi?} characterised by its strength of attraction $\alpha$. More precisely, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to patch $i$ with population $P_i$, all $N$ units being added independently and in parallel. At the end of each time step, this growth process is smoothed $n$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its extended neighborhood (8 surrounding patches). To avoid border effects such as a reflexion on the border of the world, border patches diffuse to the outside\comment[FL]{comment est-ce qu'on contrôle la perte de population induite?}. The model stops when a fixed total population \comment{n?} is reached. Grids are thus generated from the combination of the values of these four meta-parameters $\alpha$, $\beta$, $n$ and $N$, in addition to the random seed. To ease our exploration, only the distribution of density is allowed to vary rather than the size of the grid, which we fix to a 50x50 square environment of 100,000 units (cf. figure \ref{fig:spatialGen}). Typical value ranges for the other parameters are $\alpha\in\left[0.5,4.0\right]$, $\beta \in\left[0,0.3\right] $, $N\in \left[100,10000\right]$, $n\in\left[1,4\right]$.
\comment{A detailler en incluant les equations de morphogenese}[(JR) j'ai détaillé le processus, c'est suffisant ?]\comment[MLT]{Il me semble que l'on pourrait mieux expliciter la relation N/n (et peut-être du coup inverser N et n). Plus il me semble que l'on devrait dire plus clairement à quoi se réfère alpha : à une cellule $i$? au pas de temps $t$? Pourquoi ne pas indexer alpha? Enfin, je crois que l'on gagnerait encore plus en clarté si on expliquait formellement comment tous les paramètres sont reliés entre eux, notamment pour expliquer l'arrêt de la procédure d'attribution d'une nouvelle pop à la grille.} \comment[FL]{je crois que N c'est l'incrément de pop et n le nombre de fois où c'est smoothé mais c'est quoi la pop totale?}

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Including a (typical) variety of initial conditions in the sensitivity analysis}

In order to generate density grids which correspond to empirical density distributions, we select among the generated grids using an objective function which matches the point cloud of 110 metropolitan areas in Europe described by four dimensions of spatial structure : their concentration index, hierarchy index, centrality index and continuity index (cf. \cite{LeNechet2015}). A stochastic exploration of a Latin Hypercube Sampling of 2000 points in the 4-dimensional space of parameters {$\alpha$, $\beta$, $n$, $N$} gives a subset of 170 interesting grids matching empirical densities, which constituted our set of different initial spatial conditions. These are further clustered into three classes of morphology: compact (e.g. Vienna), polycentric (Liege) and discontinuous (Augsburg) in order to evaluate the non-trivial effects of urban form on simulation results. We select 15 grids of each type to work with.

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Comparing Phase Diagrams}

In order to test for the influence of spatial initial conditions on model outputs, we need a systematic method to compare phase diagrams. Indeed, we have as many phase diagrams than we have spatial grids, what makes a qualitative visual comparison not realistic. A solution is to use systematic quantitative procedures. To our knowledge there exist no well established method to compare phase diagrams in agent-based modeling and geosimulation literature. Several potential methods from other fields such as environmental science could be used, but we keep it simple and such methodological considerations are furthermore auxiliary to the main purpose of this paper. We propose therefore an intuitive measure corresponding to the share of between-diagrams variability relative to their internal variability. More formally, the distance is given by

>>>>

\corrauth{\email{juste.raimbault@polytechnique.edu}}

@@

\begin{abstract}
%\noindent
%\setlength{\parindent}{0pt}
Although simulation models of geographical systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain until their results are proven robust and representative of 'real-world' conditions. Sensitivity analysis usually includes the analysis of the effect of stochasticity on the variability of results, as well as the effects of small parameter changes. However, initial spatial conditions are usually taken for granted in geographical models, thus leaving completely unexplored the effect of spatial arrangements on the interactions of agents and of that of agents with their environment. In this paper, we present a method to assess the effect of initial spatial conditions on simulation models, using a systematic generator controlled by meta-parameters in order to create the density grids with which spatial simulation models are initialised. We show, with the example of two very classical agent-based models (Schelling's models of segregation and Sugarscape's model of resource harvesting) and a straightforward open-source work-flow using high performance computing, that the effect of space is significant in the two types of simulation. Furthermore, this effect is sometimes larger than the effect of parameters' value change. 
\end{abstract}

@@

% **************  MAIN BODY OF THE PAPER **************
\newpage

@@

Simulation within computer models has been recognised and increasingly used by geographers as an efficient tool to explore geographical processes, hypotheses and predictive scenarios within virtual laboratories \citep{batty1971modelling, batty2007model, carley1999generating, Quesneletal2009}. \comment{+ Seb \& James Doran in the 1970s??} Simulation also appears as a way to overcome the difficult analytic resolution of many spatial models which were developed in the past, as well as to explore the possible (alternative) trajectories of social and ecological systems in space and time. The specificity of geographical models compared to other social sciences' models is that space and spatial interactions are given a prime role, as geographers are driven by an explicit interest in studying the way space influences the outcomes of the model. Indeed, they usually seek to understand and model the role of space in social interactions and environmental processes, in a city for example. We think simulation can become a very good tool to achieve this, provided that models do include relevant spatial description and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of output variations to the way space is modelled. Unfortunately, the first condition is not always met, and the second is seldom even mentioned. This paper aims to fill a methodological and conceptual gap, which is a systematic testing of the sensitivity of a model's outcomes to its initial spatial conditions. To demonstrate the genericity of our approach, we develop two applications with classic simulation models commonly used as case studies for comparing and aligning simulation models \citep{Axtelletal1996, wilensky2007making}: \citet{schelling1971dynamic}'s model of segregation and \citet{EpsteinAxtell1996}'s Sugarscape model, both defined at the intra-urban level.\\

An urban system can be defined as a large number of diverse social agents interacting with one another, within a limited portion of space. Social agents thus constitute the microscopic level of urban systems, which are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects. Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000}, and emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}. Complexity is thus partially due to bifurcations, which are determinant in spatial systems~\citep{Wilson1981, Wilson2002}. Indeed, in spatially explicit simulation models, the non-linearity of local interactions is very likely to sublimate small perturbations in the initial spatial setting, making it difficult to interpret the resulting global structures. \\
Another crucial aspect of spatio-temporal complex systems in general, and cities in particular, is their non-ergodicity~\citep{pumain2012urban}, which means that cross-sectional samples in space (the different cities of a country at time t) are not equivalent to samples in time (the same city at time t, t+1, t+2...) to compute statistics such as average. On the contrary, strong spatio-temporal path-dependencies in the trajectory of individual cities and changing social environments over time provide the use of ergodic models. Similarly to what \citet{gell1995quark} calls \emph{frozen accidents} in complex system generally, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system. \\

Although this may seem obvious, cities are not regular grids, and the distribution of density (of jobs, residents, buildings, etc.) in cities is far from isotropic, even in the most sprawled cities. On the contrary, there is a significant diversity in the way people, activities and structures are distributed in cities. In Europe for example, \citet{LeNechet2015} quantifies and classifies six broad types of residential density distributions. However, most geographical models, especially cellular automata, represent cities as uniform grids of isotropic densities. Even in applied cases when GIS geometries of a particular city are used, the spatial distribution of agents is still approximated by a constant density \citep{arribas2014diverse}, although previous research shows that it is computationally and methodologically feasible to use accurate locations in a simple model such as Schelling's \citep{benenson2002entity}. The isotropic simplification is potentially harmful to the representation of the urban processes under inquiry because density and accessibility have environmental, economic and social consequences. Additionally, we expect the initial spatial distribution of agents to influence simulation results in the long run \citep{Castellanoetal2009}, because the agents' rule of action itself may depend on the spatial structure of the environment. For example, households can have different preferences with respect to the built-environment they want to live in \citep{SpielmanHarrison2014}, or agents can move around and sense their environment up to a certain distance, so that the resulting number of sensed objects will depend on the topology \citep{Banos2012} and distribution of density of the sensed environment \citep{LauriJaggi2003, FossettDietrich2009}. The way modellers abstract space is therefore a central element of geographical simulation models. A meaningful way of abstracting space might thus be to consider, not necessarily the peculiarities of every city, but at least their broad density structures.

@@

In this paper, we suggest a way of performing a sensitivity analysis to the spatial conditions of models by using a spatial generator to produce a variety of density grids, which are then fed into simulation models at initialisation. The generator being controlled by meta-parameters, we can then relate the parameters used to generate initial spatial conditions to the variation in outcome of the simulations. The purpose is two-fold: 1/ to test the robustness of simulation results to small variations of meta-parameters and 2/ to study the non-trivial effects of typical categories of spatial distribution (monocentric \textit{vs.} polycentric for example) on the results of a given model. Our approach allows for a systematic comparison of several aspects of the spatial configuration problem, which has been suggested by \citet{filatova2013spatial}, but hardly implemented and achieved in previous studies, which tend to focus on single aspects of the debate. In particular, it is applied to the effects of urban form on simulation results, using Schelling's model as a first case study and Sugarscape as a second one.

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Previous considerations on the effects of the urban form in simulation models}

%\comment[FL]{Je pense que cette partie est vraiment utile, et devrait être étoffée, en procédant peut être à une typologie plus systématique des "initial condition" qui nous intéressent. Il y a déjà une différence entre l'hétérogénéité de l'espace (densités réparties mono ou poly, etc) et les différentes façons de représenter l'espace (grilles carrées, hexagones). Par ailleurs aussi des discussions sur l'incertitude de l'information (est ce que par exemple les emplois downscalés au carreau INSEE par Pivano 2016 c'est vraiment fiable ou très ?) et la précision de la grille}

Different spatial configurations produce different distributions of income, carbon emissions, educational outcomes, etc. For example, \citet{wheeler2006urban} shows that, in the US, sprawling cities tend to be more unequal than their compact counterparts with respects to income. Dynamically, sprawl consists in the addition of new developments which have been occupied by different groups of population, resulting in a concentration of the wealthy in suburban pockets and of pockets of poverty in the inner city area \citep{jargowsky2002sprawl}. Similarly, in terms of pollution for example, \citet[p.173]{schwanen2001travel} show that{\it "deconcentration of urban land uses encourages driving and discourages the use of public transport as well as cycling and walking"}. The discussion of these effects has also been extensive in the field of geostatistics, meanly since the exposure of the Modifiable Areal Unit Problem (MAUP) \citep{Openshaw1984, FotheringhamWong1991}. Recently, \citet{Kwan2012} has argued for a careful examination of what she coins the uncertain geographic context problem (UGCoP), i.e. of the spatial configuration of geographical units even when the size and delineation of the area are the same.\\

Considerations of such issues in the geographical modelling literature are rather scarce, although there have been some noticeable attempts to evaluate the effects of the same set of mechanisms run from different initial spatial conditions. Mainly, they have focused on the impact of:
\begin{itemize}
\item The accuraty of geo-localised input data;
\item The spatial system shape, precision and boundaries;
\item Spatial heterogeneity.
\end{itemize}
We detail these approaches in the paragraphs below. Other approaches, not included in the scope of this article, include the effects of the topology of space on the processes (cf. \citet{moreno2009integrating} who explore the role of the topology of possible interactions in a Schelling model). Generally, sensitivity analyses focus either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.); either on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely both at the same time.

\paragraph{Geo-localised input data accuracy:} \citet{Thomasetal2017} show that data selection in LUTI model is inter-related with the delineation of the spatial system boundaries and scale of analysis. They provide a few examples of how the use of Exploratory Spatial Data Analysis (ESDA) prior to simulation runs can help avoiding measurement errors of model behaviour and outcomes. In the context of spatial interaction models, \citet{hagen2012new} acknowledge the dilemma between spatial resolution and computational burden, and suggest a method of adaptive zoning (where the size of destination zones depends on the distance to origin) to solve it.

\paragraph{Spatial system shape, precision, and boundaries:} \citet{Axtelletal1996} highlight the sensitivity of the average number of stable cultural regions generated to the effect of the territory width implemented in a version of the Sugarscape model which is docked (i.e. made equivalent to) to Axelrod Culture Model. \citet{FlacheHegselmann2001} show that chances for random emergence of a stable cluster of similar agents in a Schelling-like model are higher in a rectangular grid and lower in a hexagonal grid and that an irregular (Voronoi-diagram) city lattice structure favours migration stabilisation around decentralised clusters of similar agents. \citet{Banos2012} compares the behaviour of Schelling segregation model on city lattices formalized as either grid, random, scale-free and Sierpinski networks and concludes that the presence of cliques in graph-based urban structures favor segregationist behaviours. \citet{LeTexierCaruso2017}, using a set of different theoretical spatial systems, demonstrate the impact of the regularity and aggregation levels, or centrality/periphery effects, on spatial diffusion dynamics of euro coins. 
\comment[FL]{je dois compléter\\
\cite{horritt2001effects} identified the effects of grid cell size on the behaviour of a raster flood model
\cite{vazquez2002effect} ont exploré la meilleure granularité pour la précision d'un modèle d'écoulement hydrologie et le temps de calcul (à reprendre)}

\paragraph{Spatial heterogeneity:} \citet{StaufferSolomon2007} introduce asymetric interactions and empty residences in Schelling's model run on a large and regular lattice. They reveal conjoint and non-linear effects on the vacancy rates and tolerance levels on segregation patterns. \citet{Gauvinetal2010} run Schelling's segregation process in an open city-lattice to study how the variations in tolerance levels, vacancy rates and city attractiveness may create lines of vacancy lots between clusters of agents. They conclude on the functional role of vacancies, which allow weakly tolerant agents to live and be satisfied in a city environment they nevertheless perceive as hostile. \citet{HatnaBenenson2012} show that their model replications run on a 50x50 torus with 2\% of empty cells were not sensitive to the initial patterns (random and fully segregated distribution of agents).

\paragraph{Combination:} \citet{Singhetal2009} show that the segregation patterns obtained by Schelling for certain tolerance values are strictly a small city phenomenon (8x8 city-lattice) and do not work for a larger spatial lattice (100x100), where segregation appears only for certain combinaisons of tolerance threshold and vacancy density values.

\comment[clem]{Ce serait cool d'avoir aussi des exemples de simu hors Schelling pour les deux dernieres sous-parties}

In this paper, we build on the contributions of these previous studies and go further by systematically measuring the impact of the city lattice on model behaviour, which has so far only been assessed through a small set of illustrative spatial structures. We illustrate the potential generalisation of our approach by running two distinctive agent-based models: Schelling's model of residential segregation and Sugarscape.

@@

In this section, we detail the method developed to analyse the sensitivity of simulation models to initial spatial conditions. The general method workflow is illustrated in Figure \ref{fig:method}. In addition to the usual protocol (upper branch in figure \ref{fig:method}, which consists of running a model $\mu$ with various values of its parameters, we introduced the lower branch in figure \ref{fig:method}. We wanted to emphasize that spatial effects do not only derive from grid size or shape effects, but indeed from the heterogeneity of relevant geographic entities (people, housing, networks, etc.) through geographical space. In models, the initial configuration of space can be either flat (most economic models\comment[FL]{c'est peut être un peu brutal?}), or heterogeneous,  based on external databases and internal modelling (generation of synthetic population for instance \citep{bhat1999activity}).
%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[htbp] \begin{center} 
\resizebox{0.9\textwidth}{!}{ 
	\includegraphics{figures/SchemaMeta_1.png}
} \caption{General workflow of our method} \label{fig:method} \end{center} \end{figure*} %
%%%%%%%%%%%%%%%%%%%%%%

In our approach, we have introduced a spatial generator (lower branch in figure \ref{fig:method}), which itself is determined by input parameters and produces sets of spatial initial conditions. 
To allow for the exploration of the relative effects of model parameters and spatial configuration, we clustered initial spatial conditions  to represent types of spaces ex-ante (for example: monocentric or polycentric density grids). The sensitivity analysis of the model is now run against $\mu$ parameters as well as spatial parameters or spatial types. It allows the sensitivity analysis to produce qualitative conclusions regarding the influence of spatial distribution on the outputs of simulation models, alongside the classic variation of parameter values.

In order to test for the influence of spatial initial conditions on model outputs, we use a systematic method to compare phase diagrams. We define a phase diagram as a vector of  outputs\comment[FL]{Juste, peut-être des choses qui sont mal dites? : par exemple, dans diagramme de phase est-ce qu'on entend pas une "réification" de catégories - des clusters par ex? Et du coup est-ce que cela ne manque pas là?}.

Comparing phase diagrams is useful because it allows us to compare the 
Indeed, we have as many phase diagrams than we have spatial grids, what makes a qualitative visual comparison not realistic\comment[FL]{donner un ordre de grandeur du nombre de simulations?}. A solution is to use systematic quantitative procedures. To our knowledge there exist no well established method to compare phase diagrams in agent-based modeling and geosimulation literature. We used a distance between two phase diagram corresponding to the share of between-diagrams variability relative to their internal variability\comment[FL]{je ne comprends pas ce choix : pourquoi faut il considérer que des classes ayant une forte cohérence interne seraient plus éloignées? ou alors la question ne se pose pas en ces termes... Par ailleurs est ce que la formule est bien homogène du point de vue des dimensions? En tous cas je pense qu'on peut se passer de la formule à ce stade, éventuellement à mettre dans l'affichage des résultats si on en a besoin.}

%Several potential methods from other fields such as environmental science could be used, but we keep it simple and such methodological considerations are furthermore auxiliary to the main purpose of this paper. We propose therefore an intuitive measure corresponding to the share of between-diagrams variability relative to their internal variability. More formally, the distance is given by

%\begin{equation}\label{eq:phase-distance}
%d_r\left(\alpha_1,\alpha_2\right) = 2 \cdot \frac{d(f_{\vec{\alpha_1}},f_{\vec{\alpha_2}})^2}%{Var\left[f_{\vec{\alpha_1}}\right] + Var\left[f_{\vec{\alpha_2}}\right]}
%\end{equation}

%where $f_{\vec{\alpha}}(\vec{x})$ is the phase diagrams on parameters $\vec{x}$ at given meta-parameters $\vec{\alpha}$. $d$ is a functional distance that we take simply as Euclidian distance. The variances are estimated within each phase diagram $f_{\vec{\alpha_i}}$.

The last methodological point which we need to emphasis is the relation between the workflow we introduce and model exploration workflows. The ideas of multi-modeling and extensive model exploration are nothing from new as ~\cite{openshaw1983data} already advocated for ``model-crunching'', but their effective use only begins to emerge thanks to the apparition of new methods and tools together with an explosion of computation capabilities. The model exploration platform OpenMole~\citep{reuillon2013openmole} allows to embed any model as a blackbox, write modulable exploration workflow using advanced methodologies such as genetic algorithms and distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids. In our case, the workflow tool is a powerful way to embed both the sensitivity analysis and the meta-sensitivity analysis, and allow to couple any generator with any model in a straightforward way as soon as the model can take its spatial configuration as input or from an input file. In this paper, we use the OpenMole platform for spatial environment and model coupling, placing ourselves in the renewed framework of multi-modelling claimed by \citet{cottineau2015modular}.
%%%%%%%%%%%%%%%%%%%%%%
\subsection{Constructing a spatial generator of density grids}

Our spatial generator applies an urban morphogenesis model (\cite{Batty2007}), which has been generalised, explored and calibrated (\cite{Raimbault2014}). In a nutshell, grids are generated through an iterative process which, starting from a void grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment \comment[FL]{à quoi?} characterised by its strength of attraction $\alpha$. More precisely, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to patch $i$ with population $P_i$, all $N$ units being added independently and in parallel. At the end of each time step, this growth process is smoothed $n$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its extended neighborhood (8 surrounding patches). To avoid border effects such as a reflexion on the border of the world, border patches diffuse to the outside. The procedure stops when a fixed number of steps $k$ is reached. The grid then has a population of $k \times N$ (the population lost due to diffusion process to the outside is reallowed through a normalization procedure at the end of the steps). Grids are thus generated from the combination of the values of these four meta-parameters $\alpha$, $\beta$, $n$ and $N$, in addition to the random seed. To ease our exploration, only the distribution of density is allowed to vary rather than the size of the grid, which we fix to a 50x50 square environment of 100,000 units (cf. figure \ref{fig:spatialGen}). Typical value ranges for the other parameters are $\alpha\in\left[0.5,4.0\right]$, $\beta \in\left[0,0.3\right] $, $N\in \left[100,10000\right]$, $n\in\left[1,4\right]$.
\comment{A detailler en incluant les equations de morphogenese}[(JR) j'ai détaillé le processus, c'est suffisant ?]\comment[MLT]{Il me semble que l'on pourrait mieux expliciter la relation N/n (et peut-être du coup inverser N et n). Plus il me semble que l'on devrait dire plus clairement à quoi se réfère alpha : à une cellule $i$? au pas de temps $t$? Pourquoi ne pas indexer alpha? Enfin, je crois que l'on gagnerait encore plus en clarté si on expliquait formellement comment tous les paramètres sont reliés entre eux, notamment pour expliquer l'arrêt de la procédure d'attribution d'une nouvelle pop à la grille.} \comment[FL]{je crois que N c'est l'incrément de pop et n le nombre de fois où c'est smoothé mais c'est quoi la pop totale?}

@@

%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[htbp] \begin{center} 
\resizebox{0.9\textwidth}{!}{ 
	\includegraphics{figures/GridPresentation.png}
} \caption{Correspondence between European urban density structures and grids produced with the spatial generator.} 
\label{fig:densityTypes} \end{center} \end{figure*} %
%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Including a (typical) variety of initial conditions in the sensitivity analysis}

In order to generate density grids which correspond to empirical density distributions, we select among the generated grids using an objective function which matches the point cloud of 110 metropolitan areas in Europe described by four dimensions of spatial structure : their concentration index, hierarchy index, centrality index and continuity index (cf. \cite{LeNechet2015}). A stochastic exploration of a Latin Hypercube Sampling\comment[FL]{ce n'est pas explicite} of 2000 points in the 4-dimensional space of parameters {$\alpha$, $\beta$, $n$, $N$} gives a subset of 170 interesting grids matching empirical densities, which constituted our set of different initial spatial conditions. These are further clustered into three classes of morphology: compact (e.g. Vienna), polycentric (Liege) and discontinuous (Augsburg) in order to evaluate the non-trivial effects of urban form on simulation results. We select 15 grids of each type to work with in the computation of sensitivity analysis. \comment[FL]{pourquoi 15?} \comment[FL]{dire explicitement ce qu'on va faire avec ces catégories dans la partie résultats?}

@@

%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Comparing Phase Diagrams}
\comment[FL]{j'ai proposé de mettre cette partie au début de method et de la simplifier}

%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Model Exploration Workflows as an meta-sensitivity analysis method}
\comment[FL]{item}

In the following section, we briefly recall the main components of the two ``classical'' agent-based simulation models used to test how spatial density variations may impact simulation models behaviour and results, and how general the method proposed is.

\subsection{Case study models}
